A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-11-22 doc/paper1-submission/paper.tex

%% LyX 2.2.0 created this file.  For more info, see http:____www.lyx.org__.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{paper}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{xcolor}
\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__abcd}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__bcde}

@@

\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln1}

}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures__toy_ambig__soln2}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__abc}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__acd}

}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__bcd}

@@

\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s12}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s13}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures__toy_pairwise_compat__s23}

@@

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Figure
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.75\paperheight]{paper-pipeline-tools.pdf}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}

@@

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input. In version 5.0 of the Open Tree
of Life synthetic tree, this version of the taxonomic tree contained
only 41,226 leaves (while the flag-cleaned version of OTT had 2,424,255
leaves).
\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__cleaned_phylo__ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures__cleaned_ott__cleaned_ott\lyxdot tre}

@@

\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures__exemplified_phylo__ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures__exemplified_phylo_taxonomy_uncontested\lyxdot tre}

@@

\begin{figure}
  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__14__3\lyxdot tre}

@@

  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures__subproblems__17__3\lyxdot tre}

@@

\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
A trivial statemen from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__ab1c}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__acb2}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__tax}

@@

\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__opt}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures__uncontested_worsens_score__returned}

@@

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott14\lyxdot tre}

}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures__subproblem_solutions__ott17\lyxdot tre}

@@

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures__grafted_solution__grafted_solution_ottnames\lyxdot tre}
\par\end{centering}
\caption{Grafted solution. The nodes E and F are monotypic here, but will end
up being polytypic after unpruning is performed.}
\end{figure}

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.4]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures__labelled_supertree__labelled_supertree_only_taxon_names\lyxdot tre}
\par\end{centering}
\caption{Unpruned tree with internal taxa. Nodes E2, F2, and D have been re-added
to the tree. The nodes E and F are no longer monotypic.}
\end{figure}

@@

\begin{figure}
\subfloat[Input tree $T_{1}$ named ``tree1'']{%
\begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__input1\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Summary Tree $\protect\summaryTree$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__summary\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Induced Tree $\protect\summaryTree(1)$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example__induced1_color}%
\end{minipage}

>>>>

%% LyX 2.2.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{paper}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{xcolor}
\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/abcd}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/bcde}

@@

\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/soln1}

}\hfill{}\subfloat[Summary tree]{\includegraphics[width=0.45\textwidth]{Figures/toy_ambig/soln2}

@@

\begin{figure}
\subfloat[Input tree 1]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/abc}

}\hfill{}\subfloat[Input tree 2]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/acd}

}\hfill{}\subfloat[Input tree 3]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/bcd}

@@

\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s12}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s13}

}\hfill{}\subfloat[Summary tree]{\includegraphics[scale=0.6]{Figures/toy_pairwise_compat/s23}

@@

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Figure
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.75\paperheight,bb = 0 0 200 100, draft, type=eps]{paper-pipeline-tools.png}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}

@@

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input. In version 5.0 of the Open Tree
of Life synthetic tree, this version of the taxonomic tree contained
only 41,226 leaves (while the flag-cleaned version of OTT had 2,424,255
leaves).
\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures/cleaned_phylo/ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures/cleaned_phylo/ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Taxonomy Tree]{\includegraphics[scale=0.4]{Figures/cleaned_ott/cleaned_ott\lyxdot tre}

@@

\begin{figure}
\subfloat[Tree 1]{\includegraphics[scale=0.4]{Figures/exemplified_phylo/ex_2@tree1\lyxdot tre}

}\hfill{}\subfloat[Tree 2]{\includegraphics[scale=0.4]{Figures/exemplified_phylo/ex_2@tree2\lyxdot tre}

}\hfill{}\subfloat[Pruned taxonomy tree $\protect\prunedTaxonomy$]{\includegraphics[scale=0.4]{Figures/exemplified_phylo_taxonomy_uncontested\lyxdot tre}

@@

\begin{figure}
  \subfloat[Subproblem \emph{ABCD}]{\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/14/3\lyxdot tre}

@@

  \subfloat[Subproblem \emph{root}]{\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/1\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/2\lyxdot tre}%
    \hfill{}\includegraphics[width=0.3\textwidth]{Figures/subproblems/17/3\lyxdot tre}

@@

\caption{Subproblems generated from the exemplified trees shown in figure \ref{fig:exemplify-example}.
A trivial statemen from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/ab1c}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/acb2}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/tax}

@@

\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/opt}

}\hfill{}\subfloat[]{\includegraphics[scale=0.6]{Figures/uncontested_worsens_score/returned}

@@

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.45]{Figures/subproblem_solutions/ott14\lyxdot tre}

}\hfill{}\subfloat[Solution to subproblem \emph{root}]{\includegraphics[scale=0.45]{Figures/subproblem_solutions/ott17\lyxdot tre}

@@

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures/grafted_solution/grafted_solution_ottnames\lyxdot tre}
\par\end{centering}
\caption{Grafted solution. The nodes E and F are monotypic here, but will end
up being polytypic after unpruning is performed.}
\end{figure}

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3/cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.4]{example3/grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.4]{example3/unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.4]{example3/unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{Figures/labelled_supertree/labelled_supertree_only_taxon_names\lyxdot tre}
\par\end{centering}
\caption{Unpruned tree with internal taxa. Nodes E2, F2, and D have been re-added
to the tree. The nodes E and F are no longer monotypic.}
\end{figure}

@@

\begin{figure}
\subfloat[Input tree $T_{1}$ named ``tree1'']{%
\begin{minipage}[c][1\totalheight][t]{0.25\columnwidth}%
\includegraphics[scale=0.5]{annotation_example/input1\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Summary Tree $\protect\summaryTree$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example/summary\lyxdot tre}%
\end{minipage}

}\hfill{}\subfloat[Induced Tree $\protect\summaryTree(1)$]{%
\begin{minipage}[c][1\totalheight][t]{0.35\columnwidth}%
\includegraphics[scale=0.5]{annotation_example/induced1_color}%
\end{minipage}
